---
title: "Part A: Assembling the Data"
author: "Ellice Huang"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stargazer)
library(dplyr)
library(ggplot2)
```

## Import and prepare school data
```{r}
# import school data
hd2010 <- read.csv('Data/schools/hd2010.csv', header=T, stringsAsFactors = F)
hd2010$year <- 2010
hd2011 <- read.csv('Data/schools/hd2011.csv', header=T, stringsAsFactors = F)
hd2011$year <- 2011
hd2012 <- read.csv('Data/schools/hd2012.csv', header=T, stringsAsFactors = F)
hd2012$year <- 2012
hd2013 <- read.csv('Data/schools/hd2013.csv', header=T, stringsAsFactors = F)
hd2013$year <- 2013
hd2014 <- read.csv('Data/schools/hd2014.csv', header=T, stringsAsFactors = F)
hd2014$year <- 2014
hd2015 <- read.csv('Data/schools/hd2015.csv', header=T, stringsAsFactors = F)
hd2015$year <- 2015

# vertically combine all school datasets and check it was combined correctly
schools <- rbind(hd2010,hd2011,hd2012,hd2013,hd2014,hd2015)
summary(schools)
```

## Import and prepare student data
```{r}
# import data
sfa <- read.csv('Data/students/sfa1015.csv', header=T, stringsAsFactors = F)

# separate dataset by year using header labels
sfa2010 <- sfa[1:8]
sfa2010$year <- 2010
sfa2011 <- sfa[c(1,9:15)]
sfa2011$year <- 2011
sfa2012 <- sfa[c(1,16:22)]
sfa2012$year <- 2012
sfa2013 <- sfa[c(1,23:29)]
sfa2013$year <- 2013
sfa2014 <- sfa[c(1,30:36)]
sfa2014$year <- 2014
sfa2015 <- sfa[c(1,37:43)]
sfa2015$year <- 2015

# rename columns to match for vertical concatenation
sfacolumns <- c('unitid', 'scugrad', 'scugffn', 'scugffp', 'fgrnt_p', 'fgrnt_a', 'sgrnt_p', 'sgrnt_a','year')
sfa_list <- list(sfa2010,sfa2011,sfa2012,sfa2013,sfa2014,sfa2015)
ChangeNames <- function(x) { # function that renames columns 
  names(x) <- sfacolumns
  return(x)
}
sfa <- lapply(sfa_list, ChangeNames )

# combine vertically all student datasets and check it was combined correctly
students <- rbind(sfa[[1]],sfa[[2]],sfa[[3]],sfa[[4]],sfa[[5]],sfa[[6]])
summary(students)
```
## Merge school and student data on institution ID and year
```{r}
data <- merge(schools, students, by=c("unitid","year"), all = TRUE) 
data <- na.omit(data) # remove incomplete rows
```

## Assign states value by merging zip code/state data
```{r}
zip_to_state <- read.csv("Data/xwalk/zip_to_stabbr_xwalk.csv", header=T, stringsAsFactors = F)
data$zip_clean <- sapply(strsplit(as.character(data$zip), "-"), "[[", 1) #split zip codes to only first 5 numbers
data <- merge(data, zip_to_state, by.x = 'zip_clean', by.y = 'zip', all.x = TRUE)
```

## Filter data 
for:

- Tennessee schools
- undergraduate institutions only (no graduate-only institutions)
- both public and private institutions
- no NA values
```{r}
data <- data[data$stabbr=='TN',]
data <- data[data$ugoffer==1,]
data <- na.omit(data)
```

## Finalize cleaned data
variables:

- ID_IPEDS: a unique identifier for each institution
- year: the four-digit academic year, where “2010-11” is coded as 2010, “2011-12” is 2011, and so on
- degree_bach: a dummy variable that identifies bachelor's degree-granting institutions
- public: a dummy variable that identifies public institutions
- enroll_ftug: the total number of first-time, full-time degree/certificate-seeking (henceforth,“FTFT”) undergraduates
- grant_state: total amount of state and local grant aid awarded to FTFT undergraduates
- grant_federal: total amount of federal grant aid awarded to FTFT undergraduates
```{r}
data$degree_bach <- ifelse(data$iclevel == 1, 1, 0)
data$public <- ifelse(data$control == 1, 1, 0)

# total grant aid = num enrolled * percent received aid * average aid
data$grant_state <- data$scugffn * (data$fgrnt_p/100) * data$fgrnt_a 
data$grant_federal <- data$scugffn * (data$sgrnt_p/100) * data$sgrnt_a

clean œ<- data[,c("unitid","year","degree_bach","public","scugffn","grant_state","grant_federal")]
names(clean) <- c("ID_IPEDS","year","degree_bach","public","enroll_ftug","grant_state","grant_federal") # rename cols

write.csv(clean,"clean.csv")
```

